---
title: "Evaluating Treatment Safety, Adherence and Pharmacokinetics in MATIK: A Comprehensive Analysis of Phase II Trial Data"
author: "Ryan Wei  UNI: rw2844"
output: 
  bookdown::pdf_document2:
    toc: no
    collapsed: no
    theme: readable
    citation_package: natbib
#bibliography: ["p8157final.bib"]
biblio-style: "apalike"
link-citations: true
geometry: "margin=1in"
fontsize: 12pt
linestretch: 1.5
header-includes:
  - \usepackage{appendix}
---

```{r load-analysis, include = F}
source(knitr::purl("P9185_project1.Rmd", quiet=TRUE))
```

# Introduction

MATIK	is	a	newly	discovered,	highly	contagious	viral	skin	disease	that	causes serious	skin	rash	and	infections	in	patients,	and	could	even	be	life	threatening.	

Several	scientific	labs	and drug	companies	are	actively	engaged	in	developing drugs	to	battle	the	emerging	epidemic.		From	all	the	biological	and	animal	studies,three	medications,	one	oral	Pill	and	two	skin	gels,	appeared	to	be	promising.		We denote	the	three	medications	by	Pill	A, Gel	B	and	Gel	C. Phase	I	trials	have	already	been	conducted	to	determine	their	maximum	tolerated	doses	among	patients.

Due	to	urgent	need,	researchers/clinicians	decide	to	conduct	a	phase	II	trial	to evaluate	the	safety	and	adherence	of	all	the	three	medications	simultaneously.	

Based	on	the	information	collected	from	the	Phase	I	trials,	and	earlier	cell	line	and	animal	studies.	The	following	doses shown in Table \@ref(tab:dose-tab)	will	be	used	in	the	trial.	

```{r dose-tab}
# Create a data frame with the table information
table_trt <- data.frame(
  Treatment = c("Pill A", "Gel B", "Gel C"),
  Description = c("200 mg, once a day for 4 weeks",
                  "1% gel concentration; three times a day on affected areas, daily use for 4 weeks",
                  "2% gel concentration; once a day on affected areas, daily use for 4 weeks")
)

# Generate the table with kable and style it with kableExtra
kable(table_trt, "latex", booktabs = TRUE, caption = "Recommended doses/administration schedule") %>%
  kable_styling(position = "center", latex_options = "hold_position") %>%
  column_spec(1, bold = TRUE)
```

# Method

## Study Design

### Patient population	and	recruitment

Approximately	180	MATIK	patients	will	be	recruited	from	a	variety	of	health	care facilities,	including	hospitals,	primary	care	clinics,	and	community-based	health services.	Recruitment	materials	will	be	approved	by	site	Institutional	Review Boards/Ethics	Committees	(IRBs/ECs).	Accrual	is	expected	to	be	completed	in approximately	6-9	months	per	site.

### Study	Regimen

Study	participants	will	be	randomized	to	one	of	six	regimen	sequences	(Sequence	1-6,	see	Table	\@ref(tab:trt-seq-tab)). Each	sequence	will	consist	of	three	4	week	periods	of	study	product	administration	followed	by	at	least	a	one-week	washout	period.	
The	duration	of	product	administration	including	the	two	washout	periods	is approximately	15	weeks.	Participants	will	receive	study	product	at the	recommended	dose,	and	be administered	in	the	order	designated	by their	randomized	sequence	(1-6).

```{r trt-seq-tab}
study_regimen <- data.frame(
  Sequence = 1:6,
  `Period 1` = c("Pill A", "Gel C", "Gel B", "Gel B", "Pill A", "Gel C"),
  `Period 2` = c("Gel B", "Pill A", "Gel C", "Pill A", "Gel C", "Gel B"),
  `Period 3` = c("Gel C", "Gel B", "Pill A", "Gel C", "Gel B", "Pill A")
)

# Generate the table with kable and kableExtra
kable(study_regimen, "latex", booktabs = TRUE, col.names = c("Sequence", "Period 1", "period 2", "Period 3"), caption = "Study Regimen") %>%
  kable_styling(latex_options = c("hold_position"), full_width = FALSE, position = "center")
```

## Study Objectives	and	Endpoints

### Primary	Objectives

Since	MATIK	is	a	chronic	condition, patients	will	reply	on	long-term	medications	to	control	their	viral	loads.	The	primary	objective	of this	trial	is	to	compare	the	safety	and	adherence	profiles	of	Pill	A,	Gel	B,	and	Gel	C, i.e.	1)	whether	the	medications	are	safe	for	patients,	2)	and	whether	patients	could	easily	adhere	to	medication	schedules	so	that	long	term	use	is feasible.

### Secondary	Objectives

The	secondary	objective	of	the	trial	is	to	have	preliminary	assessment and	comparison	of systemic	and	local	Pharmacokinetics	(PK) of	Pill	A,	Gel	B,	and	Gel	C. Additionally,	the	investigators	are	interested	to:	1)	assess	the	correlation	of	PK with	adherence	measures	and	the	occurrence	of	adverse	events	and	2)	identify demographic	factors	associated	with	product	adherence	and	whether	they	differ by	product	used	(Pill	or	gel)	or	regimen	(three	times	a	day	or	once	a	day).

## Endpoints and	their	collection schedules

### Primary Endpoints

During	each	week	of	the	4-week	trial	periods,	participants	were	followed	up	every	week	to	record	the	following	measures: 1) Safety:	The	number	of	grade	2	or	higher	adverse	events	occurred	during that	week. 2) Adherence:	The	number	of	days	(out	of	7	days)	that	patients	are	able	to take	pills	or	apply	gels	as	prescribed.	

### Secondary Endpoints

Pharmacokinetics: The	viral	loads	in	the	blood	plasma	and	affected	skin tissues	were	measured	at	the	beginning	and	at	the	end	of each	period.

### Demographics

Demographic	data	including	age,	gender	and	race	of	the	recruited	patients were	collected	at baseline.

## Statistical Analyses

### Primary Analysis

#### Analysis on Safety of the Treatments

In this trial, patients were assigned to different sequences (see Table \@ref(tab:trt-seq-tab)) of treatment regimens to examine the comparative safety and adherence profiles of the medications for long-term management of MATIK. To investigate the potential for carry-over effects, where the sequence of administered treatments could influence subsequent outcomes, we built the following generalized mixed effects model with logit link to analysis the carry-over effects:
$$\operatorname{logit}\left(\Pr(Y_{ij} = 1 \mid \mathbf{X}_{ij}, b_{i0})\right) = \beta_0 + \beta_1\cdot \text{Period}_{ij} + \beta_2\cdot\mathbf{I}(\text{Sequence 2}_{i}= 1) + \beta_3\cdot\mathbf{I}(\text{Sequence 2}_{i}=2) + b_{0i},$$
where $Y_{ij}$ is the indicator of occurrence of any adverse event within the period $j$ for patient $i$, $i = 1,\ldots,180$, $j = 1,2,3$. We used this summarized outcome for adverse event because the adverse events are rare events. We also used the same outcome for the subsequent analyses on safety of the treatments. $\text{Period}_{ij}$ is the $j$-th period number of patient $i$. $\mathbf{I}(\text{Sequence 2}_{i}=k)$ are the dummy variables for the sequence of the treatments the patient $i$ received, where the reference level is $\text{Sequence 2}_{i}= 0$, which means the patients received either ABC or BAC, $\text{Sequence 2}_{i}= 1$ means the patients received either CAB or ACB, and $\text{Sequence 2}_{i}= 2$ means the patients received either BCA or CBA. We did this translation since testing the null hypothesis of no carry-over effects, i.e., $H_0: \ \rho_A = \rho_B = \rho_C = 0$, where $\rho_k$ is the carry-over effect of treatment $k$, is the same as testing the following null hypothesis, $H_0: \ \rho_A+ \rho_B = \rho_A+ \rho_C = \rho_B + \rho_C = 0$. Random intercepts are included to account for the clustered data.  

To analyses the treatment effects on the adverse events (after excluding the possibility of the existence of carry-over effects). We used the following generalized mixed effects model with logit link to analysis the treatment effects:
$$\begin{aligned}
\text{logit}(\Pr(Y_{ij} = 1 \mid \textbf{X}_{ij}, b_{i0})) = & \ \beta_0  + \beta_1 \cdot\mathbf{I}\left(\text{Treatment}_{ij}={\text{Gel B}}\right)+ \beta_2 \cdot \mathbf{I}\left(\text{Treatment}_{ij}={\text{Gel C}}\right)\\
&+ \beta_3 \cdot \text{Period}_{ij} 
+ \beta_4\cdot\text{Age}_i \\
&+ \beta_5\cdot\mathbf{I}(\text{Gender}_i = \text{female})+ \beta_6\cdot\mathbf{I}(\text{Race}_i = \text{black})\\
&+ \beta_7\cdot\mathbf{I}(\text{Race}_i = \text{others}) + b_{i0},
\end{aligned}$$
where $\mathbf{I}(A)$ is the indicator of event $A$, $\mathbf{I}(A) = 0$ is $A$ does not happen, $\mathbf{I}(A) = 1$ is $A$ does happen. Demographic variables are included in the model to reduce variance of the estimates and control for the confounding. Age are centered for better interpretation (average age of this study is `r round(mean(baseline.dat$age),3)`).

#### Analysis on Adherence of Treatments

As in the analysis of adverse effects above, we built the following similar generalized mixed effects model with logit link to analysis the carry-over effects on the treatment adherence:
$$\begin{aligned}\operatorname{logit}\left(p_{ijk}\mid \mathbf{X}_{ijk},b_{0i}\right) &= \beta_0 + \beta_1\cdot \text{Period}_{ij} + \beta_2\cdot\text{Week}_{ijk} \\ &+ \beta_3 \cdot\mathbf{I}(\text{Sequence 2}_{i}= 1) + \beta_4\cdot\mathbf{I}(\text{Sequence 2}_{i}=2) + b_{0i},\end{aligned}$$
where $p_{ijk}$ is the probability of the $i$-th patient adhered to the treatment in week $k$ during period $j$. We assumed the observed outcome $Y_{ijk}$, which is the number of the number of days (out of 7 days) that patients are able to take pills or apply gels as prescribed, follows a $\operatorname{Binomial}(7,p_{ijk})$ distribution. We also used the same outcome for the subsequent analyses on adherence of the treatments. Random intercepts are included to account for the clustered data.  

To analyses the treatment effects on the treatment adherence (after excluding the possibility of the existence of carry-over effects). We used the following generalized mixed effects model with logit link to analysis the treatment effects:
$$\begin{aligned}
\text{logit}(p_{ijk}\mid \mathbf{X}_{ijk},b_{0i}) = & \ \beta_0  + \beta_1 \cdot\mathbf{I}\left(\text{Treatment}_{ij}={\text{Gel B}}\right)+ \beta_2 \cdot \mathbf{I}\left(\text{Treatment}_{ij}={\text{Gel C}}\right)\\
& +\beta_3 \cdot \text{Period}_{ij} + \beta_4\cdot\text{Week}_{ijk} \\
&+ \beta_5\cdot\text{Age}_i + \beta_6\cdot\mathbf{I}(\text{Gender}_i = \text{female})+ \beta_7\cdot\mathbf{I}(\text{Race}_i = \text{black}) \\
&+ \beta_8\cdot\mathbf{I}(\text{Race}_i = \text{others}) + b_{i0}.
\end{aligned}$$


### Secondary Analysis

#### Correlation of	Pharmacokinetics with	Adherence	Measures and the Occurrence	of Adverse Events

To investigate the association between safety, adherence and PK outcomes, we first built the following models to exam the existence of carry-over effect, as we did in the primary analyses:
$$\begin{aligned}\Delta\text{Skin Viral Load}_{ij} &=\beta_0 + \beta_1\cdot \text{Period}_{ij} + \beta_2\cdot\mathbf{I}(\text{Sequence 2}_{i}= 1) \\&+ \beta_3\cdot\mathbf{I}(\text{Sequence 2}_{i}=2) + b_{0i},
\end{aligned}
$$

$$
\begin{aligned}
\Delta\text{Blood Viral Load}_{ij} &=\gamma_0 + \gamma_1\cdot \text{Period}_{ij} + \gamma_3\cdot\mathbf{I}(\text{Sequence 2}_{i}= 1)\\ & + \gamma_3\cdot\mathbf{I}(\text{Sequence 2}_{i}=2) + b_{0i}.
\end{aligned}
$$

Here $\Delta\text{Skin Viral Load}_{ij}$ and $\Delta\text{Blood Viral Load}_{ij}$ are the change in skin/ blood viral load for patient $i$, calculated by subtracting after period (treatment) viral load from the viral load before each period (baseline viral load for $j = 1$, viral loads after wash out period for $j = 2,3$), respectively.

To analyses the effects of safety outcomes and adherence outcomes on PK outcomes, together with the treatment effects on systemic	and	local viral loads, the following linear mixed effects models are built:
$$\begin{aligned}\mathrm{E}[\Delta\text{Skin Viral Load}_{ij}\mid \mathbf{X}_{ij}, b_{0i}] &= \ \beta_0  + \beta_1 \cdot\mathbf{I}\left(\text{Treatment}_{ij}={\text{Gel B}}\right)\\
&+ \beta_2 \cdot \mathbf{I}\left(\text{Treatment}_{ij}={\text{Gel C}}\right) \\
& + \beta_3 \cdot \text{Period}_{ij} + \beta_4\cdot\mathbf{I}\left(\text{Number of adverse event}_{ij}\ge1\right)  \\ &+ \beta_5\cdot\text{Number of days of adherence}_{ij}\\
&+\beta_6\cdot\text{Age}_i + \beta_7\cdot\mathbf{I}(\text{Gender}_i = \text{female})\\
&+ \beta_8\cdot\mathbf{I}(\text{Race}_i = \text{black}) + \beta_9\cdot\mathbf{I}(\text{Race}_i = \text{others}) + b_{i0},
\end{aligned}
$$
and
$$\begin{aligned}\mathrm{E}[\Delta\text{Blood Viral Load}_{ij}\mid \mathbf{X}_{ij}, b_{0i}] &= \ \gamma_0  + \gamma_1 \cdot\mathbf{I}\left(\text{Treatment}_{ij}={\text{Gel B}}\right)\\
&+ \gamma_2 \cdot \mathbf{I}\left(\text{Treatment}_{ij}={\text{Gel C}}\right) \\
& + \gamma_3 \cdot \text{Period}_{ij} + \gamma_4\cdot\mathbf{I}\left(\text{Number of adverse event}_{ij}\ge1\right)  \\ &+ \gamma_5\cdot\text{Number of days of adherence}_{ij}\\
&+\gamma_6\cdot\text{Age}_i + \gamma_7\cdot\mathbf{I}(\text{Gender}_i = \text{female})\\
&+ \gamma_8\cdot\mathbf{I}(\text{Race}_i = \text{black}) + \gamma_9\cdot\mathbf{I}(\text{Race}_i = \text{others}) + b_{i0},
\end{aligned}
$$
where the number of days of adherence in the period is centered at the mean adherence day of that period.

#### Identify Demographic Factors associated with Product Adherence

In order to identify demographic factors associated	with product adherence and whether	they	differ by product used	(Pill	or Gel) or	regimen	(three times	a	day	or once	a	day), we built the following marginal models including product/regimen and their interactions with demographic variables in the models:
$$
\begin{aligned}
\operatorname{logit}\left(p_{ijk}\mid \mathbf{X}_{ijk}\right) & = \beta_0 + \beta_1\cdot \text{Period}_{ij} + \beta_2\cdot\text{Week}_{ijk}\\
&+ \beta_3\cdot\text{Age}_i + \beta_4\cdot \mathbf{I}(\text{Gender}_i =\text{female} ) + \beta_5\cdot\mathbf{I}(\text{Race}_i = \text{black}) \\
& + \beta_6\cdot\mathbf{I}(\text{Race}_i = \text{others})\\
& + \beta_7\cdot \mathbf{I}(\text{Product Type}_{ij} = \text{Gel}) + \beta_8\cdot \text{Age}_i\cdot \mathbf{I}(\text{Product Type}_{ij} = \text{Gel}) \\
& + \beta_9\cdot \mathbf{I}(\text{Gender}_i =\text{female} )\cdot \mathbf{I}(\text{Product Type}_{ij} = \text{Gel})\\
& + \beta_{10}\cdot\mathbf{I}(\text{Race}_i = \text{black})\cdot \mathbf{I}(\text{Product Type}_{ij} = \text{Gel})\\
&+ \beta_{11}\cdot\mathbf{I}(\text{Race}_i = \text{others})\cdot \mathbf{I}(\text{Product Type}_{ij} = \text{Gel}),
\end{aligned}
$$
and

$$
\begin{aligned}
\operatorname{logit}\left(p_{ijk}\mid \mathbf{X}_{ijk}\right) & = \gamma_0 + \gamma_1\cdot \text{Period}_{ij} + \gamma_2\cdot\text{Week}_{ijk}\\
&+ \gamma_3\cdot\text{Age}_i + \gamma_4\cdot \mathbf{I}(\text{Gender}_i =\text{female} ) + \gamma_5\cdot\mathbf{I}(\text{Race}_i = \text{black}) \\
& + \gamma_6\cdot\mathbf{I}(\text{Race}_i = \text{others})\\
& + \gamma_7\cdot \mathbf{I}(\text{Regimen}_{ij} = \text{Three times a day}) \\
& +\gamma_8\cdot \text{Age}_i\cdot \mathbf{I}(\text{Regimen}_{ij} = \text{Three times a day}) \\
& + \gamma_9\cdot \mathbf{I}(\text{Gender}_i =\text{female} )\cdot \mathbf{I}(\text{Regimen}_{ij} = \text{Three times a day})\\
& + \gamma_{10}\cdot\mathbf{I}(\text{Race}_i = \text{black})\cdot \mathbf{I}(\text{Regimen}_{ij} = \text{Three times a day})\\
&+ \gamma_{11}\cdot\mathbf{I}(\text{Race}_i = \text{others})\cdot \mathbf{I}(\text{Regimen}_{ij} = \text{Three times a day}),
\end{aligned}
$$
where age are centered in both models above.

# Result

## Baseline Characteristics of the Study Population

Table \@ref(tab:table-1) shows the distributions of the baseline covariates, including demographic variables and baseline viral loads. The similarities in these baseline measures suggest that the randomization process was successful in distributing known potential confounders evenly across groups. Such balance is indicative of the effectiveness of the randomization procedure, enhancing the internal validity of the study by reducing the likelihood that observed differences in outcomes can be attributed to pre-existing differences among treatment groups. This equitable distribution of baseline characteristics establishes a solid foundation for subsequent analyses of the treatments.

```{r table-1}
baseline.demo <-
  left_join(Adhere.demo, baseline.dat %>% select(ptid, bviral0, sviral0)) %>% 
  left_join(
    .,
    endpoints.period %>% select(ptid, sequence) %>% distinct()
  )
library(table1)
baseline.demo$gender <- factor(baseline.demo$gender, levels = c("male", "female"), labels = c("Male", "Female"))
baseline.demo$race <- factor(baseline.demo$race, levels = c("white", "black", "others"), labels = c("White", "Black", "Others"))


label(baseline.demo$gender) = "Gender"
label(baseline.demo$race) = "Race"
label(baseline.demo$age) = "Age"
label(baseline.demo$bviral0) = "Baseline blood viral load"
label(baseline.demo$sviral0) = "Baseline skin viral load"
label(baseline.demo$sequence) = "Treatment sequence"
library(gt)
demo.tab <-
  table1(~ age + gender + race + bviral0 + sviral0 | sequence,
    data=baseline.demo, caption="Descriptive statistics of baseline characteristics of the 180 subjects by treatment sequences.") %>% t1kable(booktabs = TRUE) %>%  kable_styling(latex_options = c("hold_position", "scale_down"))
demo.tab
```

## Longitudinal Outcomes of the Patients

Figure \@ref(fig:long-plot) shows the change of primary outcomes, the safety outcome and the adherence, across the 4 weeks in one treatment period. Since adverse events are rare, we summarized the number of patients who ever experienced adverse events during a week. From the upper panel we can see that there is a increasing trend of experiencing adverse event towards the end of the period for all treatments. From the lower panel, we see a the lines are trend downwards, which means that the odds of adherence are decreasing over time.

```{r long-plot, fig.cap="Change of primary outcomes across week in a treatment period. Upper panel: Total number of subject who had experienced adverse events, stratified by treatments. Lower panel: Change of logarithm of odds of adherence, stratified by treatments, the blue line is the population trend fitted by linear model."}
library(latex2exp)
library(ggpubr)

plot.ae <- 
endpoints.AE %>% 
  filter(AE ==1) %>%
  group_by(drug, week) %>%
  ggplot(aes(x=week))+
  geom_bar()+
  facet_grid(.~factor(drug, levels=c('Pill A','Gel B','Gel C')))+
  theme_bw()+
  ylab("Number of Subjects")+
  xlab("Week Number in a Treatment Period")

#plot.ae

plot.adhere <- 
endpoints.Adhere %>% 
  # each line is adherence of a drug taken by a subject
  ggplot(aes(x = week, y = log(Adhere/7/ (1-Adhere/7)))) +
  geom_line(alpha = 0.2,aes(x = week, y = log(Adhere/7/ (1-Adhere/7)),group = interaction(ptid, drug))) + 
  geom_point(alpha = 0.1, size = 0.8,aes(x = week, y = log(Adhere/7/ (1-Adhere/7)),group = interaction(ptid, drug)))+
  geom_smooth(method = "lm")+
  facet_grid(.~drug)+
  theme_bw() +
  ylab(TeX("$\\log\\left(\\frac{p_{ijk}}{1-p_{ijk}}\\right)$"))

#plot.adhere

plot.PK <-
endpoints.PK %>%
  ggplot(aes(y=dvalue, x=period))+
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) +
  geom_line(alpha = 0.2, aes(group = ptid)) + 
  geom_point(alpha = 0.1, size = 0.8) +
  facet_grid(.~measure)+
  xlab("Period")+
  ylab("Viral Load Difference")+
  theme_bw()

#plot.PK

ggarrange(plot.ae, plot.adhere, 
          #labels = c("(a)", "(b)"),
          ncol = 1, nrow = 2)
```

To identify potential demographic factors that affect the treatments adherence, and their interaction with product type and regimen, figure \@ref(fig:plot-adhere-demo) shows the adherence rate versus age stratified by product type and regimen, as well as gender and race. Overall, we found a higher adherence rate among patients who were using pill than who were using gel. We also found that patients who were prescribed a lower frequency of using pill/gel daily has a higher adherence rate. We didn't observe a clearly relationship between gender/race and adherence rate, however, we did found a increase in adherence rate as age increases.

```{r plot-adhere-demo, fig.cap="Adherence rate of a treatment period across different age, stratified by product type/regimen and demographic factors, with loess smoothing curve."}
plot.adhere.demo <- 
endpoints.PK %>%
  mutate(product_type = if_else(drug=="Pill A", "Drug", "Gel"),
         regimen = if_else(drug== "Gel B", "3 times/day", "1 time/day"),
         product_dosage = case_when(
           drug == "Pill A" ~ "Drug, 1 time/day",
           drug == "Gel B" ~ "Gel, 3 times/day",
           drug == "Gel C" ~"Gel, 1 time/day"
         )) %>%
  ggplot(aes(y = Adhere_rate, x = age))+
  #geom_boxplot()+
  geom_point()+
  geom_hline(yintercept = 1, linetype = 2, color = "grey")+
  geom_smooth(color="blue", method="loess")+
  facet_grid(gender + race~regimen + product_type)+
  theme_bw()+
  ylab("Adherence Rate of a Treatment Period")+
  xlab("Age") + 
  labs(caption = "(Age is centered)")

library(patchwork)
plot.adhere.demo
```

For the secondary endpoint, figure \@ref(fig:pk-plots) shows the overall viral loads change of different treatment groups. We found that Pill A achieved a higher viral loads reduction in both skin viral load and blood viral load compared to Gel B and Gel C. We also found that the occurrence of adverse event is associated with a slightly lower reduction in viral loads, while higher adherence is associated with a higher reduction in viral loads, especially in patients who were using Gel B and Gel C.

```{r pk-plots,fig.cap="Viral loads change of different treatment groups. Upper panel: stratified by measurement type and occurence of adverse event. Lower panel: change in viral loads versus the number of adherence days, stratified by treatments and measurement types.", fig.width=6, fig.height=8, message=F}
plot.pk.ae <-
 endpoints.PK %>%
  ggplot(aes(y=dvalue, x=drug))+
  geom_boxplot()+
  facet_grid(AE_ind ~measure)+
  theme_bw()+
  ylab("Viral Load Difference")+
  xlab("Treatments")

plot.pk.adhere <-
 endpoints.PK %>%
  ggplot(aes(y=dvalue, x=Adhere_sum))+
  #geom_boxplot(aes(group = Adhere_sum))+
  geom_point(size=0.8)+
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) +
  facet_grid(drug~measure)+
  xlab("Number of Adherence Days in a Treatment Period")+
  ylab("Viral Load Difference")+
  theme_bw()
  
ggarrange(plot.pk.ae, plot.pk.adhere,
          ncol = 1, nrow =2)
```

## Carry-over Effects

Table \@ref(tab:carryover-ae-tab)-\@ref(tab:carryover-pk-sviral-tab) shows the coefficients estimates of the fixed effects of the model analyzing carry-over effects of the treatments on the occurrence of adverse events, the adherence and the PK (blood and skin), respectively. We did not find any significant carry-over effects of treatments on both outcomes. The p-value of the test against the null hypotheses of no carry-over effects (with degree of freedoms equal 2) on safety outcome, adherence and PK (blood and skin) are $0.761$, $0.891$, $0.990$, and $0.774$, respectively, which means that we cannot reject the null of no carry-over effects at $\alpha = 0.05$ significant level in both cases.

```{r carryover-ae-tab}
tbl.ae.co.model <-
  AE.crossover.model %>% 
  tbl_regression(
    exponentiate = T,
    label = list(
      period ~ "Period",
      seq2 ~ "Sequence"
    )
  ) %>% 
  modify_header(
  update = list(p.value = "p-value", ci = "95% CI", label = "Characteristics", estimate = "OR")
  ) 


tbl.ae.co.model %>% as_kable(booktabs = T, caption = "Coefficent estimates of the carry-over effect model of adverse events.") %>%  kable_styling(latex_options = c("hold_position"))
```

```{r carryover-adhere-tab}
tbl.adhere.co.model <-
  Adhere.crossover.model %>% 
  tbl_regression(
    exponentiate = T,
    label = list(
      period ~ "Period",
      seq2 ~ "Sequence"
    )
  ) %>% 
  modify_header(
  update = list(p.value = "p-value", ci = "95% CI", label = "Characteristics", estimate = "OR")
  ) 


tbl.adhere.co.model %>%   as_kable(booktabs = T, caption = "Coefficent estimates of the carry-over effect model of adherence.") %>%  kable_styling(latex_options = c("hold_position"))
```

```{r carryover-pk-bviral-tab}
tbl.pk.co.model.b <-
  PK.crossover.model.bviral %>% 
  tbl_regression(
    exponentiate = F,
    label = list(
      period ~ "Period",
      seq2 ~ "Sequence"
    )
  ) %>% 
  modify_header(
  update = list(p.value = "p-value", ci = "$95\\%$ CI", label = "Characteristics", estimate = "$\\hat\\beta$")
  ) 


tbl.pk.co.model.b %>%   as_kable(booktabs = T, caption = "Coefficent estimates of the carry-over effect model of blood viral loads.", escape = F) %>%  kable_styling(latex_options = c("hold_position"))
```

```{r carryover-pk-sviral-tab}
tbl.pk.co.model.s <-
  PK.crossover.model.sviral %>% 
  tbl_regression(
    exponentiate = F,
    label = list(
      period ~ "Period",
      seq2 ~ "Sequence"
    )
  ) %>% 
  modify_header(
  update = list(p.value = "p-value", ci =  "$95\\%$ CI", label = "Characteristics", estimate = "$\\hat\\beta$")
  ) 


tbl.pk.co.model.s %>%   as_kable(booktabs = T, caption = "Coefficent estimates of the carry-over effect model of skin viral loads.", escape = F) %>%  kable_styling(latex_options = c("hold_position"))
```

## Safety of the Treatments

Table \@ref(tab:ae-tab) shows the coefficient estimates of the treatment effects, as well as effects of demographic factors, on the safety outcome. From the table we can see that after controlling for time variable and demographic factors, none of the treatments shows a significant effect on the occurrence of adverse event. However, we found that age has a significant effect on the occurrence of adverse event (OR = 1.22 with 95\% CI 1.12-1.32), which means that on average, an additional year of age beyond the mean age of the study population (`r round(mean(baseline.dat$age),3)` years old) is associated with 22\% increase in the odds of experiencing at least one adverse event within the treatment period.

```{r ae-tab}
model.AE %>% 
  tbl_regression(
    exponentiate = T,
    label = list(
      drug ~ "Treatment",
      period ~ "Period",
      age ~ "Age",
      gender ~ "Gender",
      race ~ "Race"
    ),
    include = everything()) %>% 
  modify_header(
  update = list(p.value = "p-value", ci = "95% CI", label = "Characteristics", estimate = "OR")
  ) %>% 
  as_kable(booktabs = T, caption = "Coefficent estimates of the treatment effects on the safety outcome.") %>%  
  kable_styling(latex_options = c("hold_position"))
```


## Adherence of the Treatments

Table \@ref(tab:adhere-tab) shows the coefficient estimates of the treatment effects, as well as effects of demographic factors, on the adherence of treatments. From the table we can see that after controlling for time variable and demographic factors, both Gel B and Gel C have significant effects on the adherence of treatments. Specifically, compared to Pill A, patients who were using Gel B has a 85\% lower odds of adherence (OR = 0.15 with 95\% CI 0.12-0.17), while patients who were using Gel C has a 79\% lower odds of adherence (OR = 0.21 with 95\% CI 0.18-0.24). Age still has a significant effect on the adherence (OR = 1.22 with 95\% CI 1.19-1.25), which means that on average, for each additional year of age beyond the mean age of the study population, the odds of adhering to the treatment is increase by 22\%.

```{r adhere-tab}
model.Adhere %>% 
  tbl_regression(
    exponentiate = T,
    label = list(
      week ~ "Week",
      drug ~ "Treatment",
      period ~ "Period",
      age ~ "Age",
      gender ~ "Gender",
      race ~ "Race"
    )) %>% 
  modify_header(
  update = list(p.value = "p-value", ci = "95% CI", label = "Characteristics", estimate = "OR")
  ) %>% 
  as_kable(booktabs = T, caption = "Coefficent estimates of the treatment effects on the adherence.") %>%  
  kable_styling(latex_options = c("hold_position"))
```

## Assessment	and	Comparison Pharmacokinetics of the Treatments

Table \@ref(tab:pk-tab) shows the coefficients estimates from the two models analyzing the effects of treatments, occurrence of adverse events and adherence on the systemic (blood) and	local (skin) pharmacokinetics. From the results we can see that on average, Gel B and Gel C have a lower reduction on both skin and blood viral loads, compared with Pill A, and these effects are significant except for Gel B on blood viral loads. We also found that for each additional day of adherence beyond the average within a period, the viral load reduction increased by 1.6 in blood and 2.3 in skin, which are significant effects. The occurrence of adverse events within the period decrease the  viral loads reduction by 23 and 54 in blood and skin, respectively, which are also significant effects. For demographic factors, we found that age is significantly negatively associated with viral loads reduction.

```{r pk-tab}
tab.bviral <-
model.PK.bviral %>% 
  tbl_regression(
    exponentiate = F,
    label = list(
      Adhere_sum_centered ~ "Days of adherece (within a period, centered)",
      drug ~ "Treatment",
      period ~ "Period",
      age ~ "Age",
      gender ~ "Gender",
      race ~ "Race",
      AE_ind ~ "Occurence of adverse effects"
    )) %>% 
  modify_header(
  update = list(p.value = "p-value", ci = "$95\\%$ CI", label = "Characteristics", estimate = "$\\hat\\beta$")
  )
#tab.bviral

tab.sviral <-
model.PK.sviral %>% 
  tbl_regression(
    exponentiate = F,
    label = list(
      Adhere_sum_centered ~ "Days of adherece (within a period, centered)",
      drug ~ "Treatment",
      period ~ "Period",
      age ~ "Age",
      gender ~ "Gender",
      race ~ "Race",
      AE_ind ~ "Occurence of adverse effects"
    )) %>% 
  modify_header(
  update = list(p.value = "p-value", ci = "$95\\%$ CI", label = "Characteristics", estimate = "$\\hat\\gamma$")
  )
#tab.sviral

tbl_merge(
  list(tab.bviral, tab.sviral),
) %>% 
  as_kable(booktabs = T, caption = "Coefficent estimates of the treatment effects on the pharmacokenitics.", escape = F) %>%  kable_styling(latex_options = c("hold_position", "scale_down")) %>% 
  add_header_above(c("", "Blood viral loads model" = 2, "Skin viral loads model" = 2))
```


## Demographic Factors associated with Adherence

Table \@ref(tab:adhere-demo-tab) shows the coefficient estimates of the models investigating demographic factors associated with adherence and the effects of product type and regimen on adherence. In both models, age is associated with a increase in the odds of adherence and week is associated with a decrease in the odds of adherence, which are in align with the results from the primary analyses. For product type, the odds of adherence for patients using gels, on average, is 77\% lower than patients using pills, no significant interactions with demographic factors are found. For treatment regimen, patients who are prescribed take treatments three times a day has a 51\% lower odds of adherence than patients who are prescribed to take treatments once a day. Age has a significant interaction with treatment regimen.

```{r adhere-demo-tab}
adhere.demo.tab.product <-
  model.Adhere.product.1 %>% 
  tbl_regression(
    exponentiate = T,
    label = list(
      product ~ "Product type",
      period ~ "Period",
      age ~ "Age",
      gender ~ "Gender",
      race ~ "Race",
      week~ "Week",
      `age:product` ~ "Age $\\times$ Product type",
      `product:gender` ~ "Product type $\\times$ Gender",
      `product:race` ~ "Product type $\\times$ Race"
    )
  ) %>% 
  modify_header(
  update = list(p.value = "p-value", ci = "$95\\%$ CI", label = "Characteristics", estimate = "OR")
  )

adhere.demo.tab.freq <-
  model.Adhere.freq.1 %>% 
  tbl_regression(
    exponentiate = T,
    label = list(
      freq ~ "Regimen",
      period ~ "Period",
      age ~ "Age",
      gender ~ "Gender",
      race ~ "Race",
      week~ "Week",
      `age:freq` ~ "Age $\\times$ Regimen",
      `freq:gender` ~ "Regimen $\\times$ Gender",
      `freq:race` ~ "Regimen $\\times$ Race"
    )
  ) %>% 
  modify_header(
  update = list(p.value = "p-value", ci = "$95\\%$ CI", label = "Characteristics", estimate = "OR")
  )

tbl_merge(list(adhere.demo.tab.product, adhere.demo.tab.freq)) %>% 
  as_kable(booktabs = T, caption = "Coefficent estimates of the effects of demographic factors and product used on the adherence.", escape = F) %>%  kable_styling(latex_options = c("hold_position", "scale_down")) %>% 
  add_header_above(c("", "Model for product type" = 3, "Model for regimen" = 3))
```

# Discussion

The analysis of the MATIK Phase II trial data provides a robust foundation for advancing to Phase III. In this discussion, we synthesize the results and delineate the implications for future trials, particularly in the context of treatment efficacy, safety, and patient adherence.

This trial's balanced baseline characteristics across treatment groups reinforce the effectiveness of the randomization procedure, which is crucial for the internal validity of the study. This balance minimizes confounding factors, thereby enabling a more accurate assessment of the treatments' true effects. Notably, the trend towards increased adverse events over time necessitates close monitoring in subsequent studies.

In considering treatments for a Phase III trial, the data suggests that Pill A exhibits a superior adherence profile compared to Gels B and C while not increasing the risk of experiencing adverse events. Given the importance of adherence in skin disease conditions management, Pill A emerges as a strong candidate for further investigation. The significant viral load reductions observed with Pill A warrant special attention, as they highlight its robust therapeutic efficacy.

For the Phase III trial design, a randomized controlled trial (RCT) remains the gold standard to confirm these findings. The RCT should be adequately powered to detect differences in adherence and viral load reductions, taking into account the observed effects from Phase II. Additionally, stratification by demographic factors, as indicated by the increased adherence with age, could be beneficial. Information from this trial that will be valuable for planning Phase III includes the adherence rates over time, the incidence and pattern of adverse events, and the demographic factors influencing treatment outcomes. Additional data that would be beneficial includes long-term follow-up for safety and efficacy, quality of life assessments, and pharmacoeconomic analyses to inform the broader impact of the treatments. To calculate the sample size needed for the Phase III study, the investigator needs to specify type I, II error rates, and the effect sizes.

This analysis is not without limitations. The generalizability of the findings is constrained by the sample size and the specific patient population studied. Furthermore, the occurrence of adverse events, though monitored, may not fully encapsulate the long-term safety profile of the treatments. Adherence could be self-reported, which can introduce reporting bias. Phase III trials should aim to include objective adherence measures and a more diverse patient population to enhance the external validity of the findings.

In conclusion, the Phase II trial findings for MATIK treatment present a compelling case for advancing Pill A into Phase III trials, emphasizing its adherence advantage and significant viral load reduction without increasing adverse events. As we proceed, it will be critical to address the limitations observed, particularly by expanding the sample size and incorporating objective adherence measures. The upcoming Phase III trial will not only seek to corroborate these promising results but also expand our understanding of Pill A's long-term safety and efficacy, ultimately aiming to improve patient care in skin disease management


\newpage
# Appendix {-}

## A.1 Code for This Report {-}

\fontsize{10pt}{12pt}\selectfont
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

