---
title: "project4_analysis"
author: "Ryan Wei"
date: "2024-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(survival)
library(KMsurv)
library(flexsurv)
library(ggsurvfit)
library(survminer)
library(patchwork)
library(lme4)
library(nlme)
library(geepack)
library(gee)
library(glmtoolbox)
```


# Study design

## Randomization Procedure

Participants will be randomly assigned to receive injections of either100 $\mu g$ of mRNA-1273 vaccine or a placebo control in a 1:1 randomization ratio. Randomization will be stratified based on age and, if they are $<$ 65 years of age, based on the presence or absence of risk factors for severe illness from COVID-19 based on CDC recommendation as of Mar 2020. There will be 3 strata for randomization: $\ge$ 65 years, $<$ 65 years and categorized to be at increased risk ("at risk") for the complications of COVID-19, and $<$ 65 years "not at risk". 

Risk will be defined based on the study participants’ relevant past and current medical history. At least 25\% of enrolled participants, but not more than 40\%, will be either $\ge$ 65 years of age or $<$ 65 years of age and "at risk" at Screening.

Participants who are less than 65 years old will be categorized as at risk for severe COVID-19 illness if they have at least 1 of the following risk factors at Screening:

- Chronic lung disease (eg, emphysema and chronic bronchitis, 
idiopathic pulmonary fibrosis, and cystic fibrosis) or 
moderate to severe asthma
- Significant cardiac disease (eg, heart failure, coronary artery 
disease, congenital heart disease, cardiomyopathies, and 
pulmonary hypertension)
- Severe obesity (body mass index ≥ 40 kg/m2
)
- Diabetes (Type 1, Type 2 or gestational)
- Liver disease
- Human Immunodeficiency Virus (HIV) infection

All participants will be assessed for efficacy and safety endpoints and provide a nasopharyngeal (NP) swab sample and blood sample before the first and second dose of IP in addition to a series of post-dose blood samples for immunogenicity through 24 months after the second dose of IP. Efficacy assessments will include surveillance for COVID-19 with RT-PCR confirmation of SARS-CoV-2 infection after the first and second dose of IP. 

## Statistical Analysis

Statistical Hypotheses: For the primary efficacy objective, the null hypothesis of this study is that the VE of mRNA-1273 to prevent first occurrence of COVID-19 is $\le 30\%$ (ie, $H_0:$
efficacy: VE $\le 0.3$).

The study will be considered to meet the primary efficacy objective if the corresponding CI of VE rules out 30\% at the primary analysis. In the primary analysis of VE of COVID-19, cases will be counted starting 14 days after the second dose of IP.

Vaccine efficacy is defined as the percent reduction in the hazard of the primary endpoint (mRNA-1273 vs placebo). Equivalently, the null hypothesis is:
- $H_0:$ efficacy: hazard ratio (HR) $\ge$ 0.7 (equivalently, proportional 
hazards VE $\le$ 0.3).

A stratified Cox proportional hazard model will be used to assess the magnitude of the treatment group difference (ie, HR) between mRNA-1273 and placebo at a 1-sided 0.025 significance level


## Sample Size Calculation

- incident rate in the vaccine group : $p_T$

- incident rate in the control group : $p_C = 0.0075$

- vaccine efficacy: $\pi = 1-\frac{p_T}{p_C} = 1-R$

- null hypothesis: $H_0:\pi \le \pi_0$

- $u = n_T/n_C$

## Method 1


- number of cases in the vaccine group follows a Poisson distribution with rate $\lambda_T = n_Tp_T$

- number of cases in the controlled group follows a Poisson distribution with rate $\lambda_C = n_Cp_C$

- number of cases in the vaccine group given the total number of cases $S$: $\operatorname{Binomial}(S,\theta)$, where $\theta = \frac{\lambda_T}{\lambda_T + \lambda_C} = \frac{1-\pi}{1-\pi + u}$

- rewrite null hypothesis $H_0:\theta \ge \theta_0$


- target risk reduction: $\pi_1 = 0.6  = 1-\frac{p_T}{p_C}$, $p_T = 0.003$, $\theta_1 = 0.2857143$

- rejection margin risk reduction: $\pi_0 = 0.3$, $\theta_0 = 0.4117647$


$$n=\frac{\left[z_\alpha \sqrt{\theta_0\left(1-\theta_0\right)}+z_\beta \sqrt{\theta(1-\theta)}\right]^2}{\left(p_T+p_C\right)\left(\theta-\theta_0\right)^2}$$

```{r}
pC = 0.0075
u = 1
pi0 = 0.3
theta0 = (1 - pi0)/(1 - pi0 + u)
pi1 = 0.6
pT = (1 - pi1) * pC
theta1 = (1 - pi1)/(1 - pi1 + u)

alpha = 0.025
beta = 0.90

sample.size.VE.low <- function(pT, pC, theta0, theta1, alpha, beta){
  z.alpha = qnorm(1-alpha)
  z.beta = qnorm(beta)
  nom = (z.alpha * sqrt(theta0 * (1 - theta0)) + z.beta * sqrt(theta1 * (1 - theta1)))^2
  denom = (pT + pC) * (theta1 - theta0)^2
  return(nom/denom)
}

sample.size.VE.low(pT,pC,theta0,theta1,alpha,beta)
```



# SAE

```{r}
long.dat <- readxl::read_excel("./Q2b.xlsx")
baseline.dat <- readxl::read_excel("./Q2b_BL.xlsx")

long.dat <- left_join(long.dat, baseline.dat, by = "ID") %>% 
  mutate(OBS = if_else(is.na(SAE), 0, 1)) %>% 
  mutate(TIME = as.factor(TIME),
         GROUP = as.factor(GROUP),
         SITE = as.factor(SITE),
         SEX = as.factor(SEX),
         SAE = as.factor(SAE))

glmm.fit <- glmer(SAE ~ TIME * GROUP + SEX + AGE + (1|ID) + (1|SITE) , data = long.dat, family = binomial)

obs.prob.fit <- glmgee(OBS ~ TIME + GROUP + AGE + GENDER + SITE, id = ID, data = long.dat, family = binomial, corr = )
```


# Survival

```{r}
survival.dat <- readxl::read_excel("./Q2c.xlsx",col_types = rep("numeric", 5)) %>% 
  mutate(InfectionTime = if_else(is.na(InfectionTime), LastFUTime, InfectionTime))

km.fit <- survfit(Surv(EnrollmentTime, InfectionTime, Infection) ~ 1, data = survival.dat)

ggsurvplot(km.fit, data = survival.dat, pval.method = TRUE, conf.int = TRUE, censor = F, risk.table = T,tables.height = 0.3)

print(km.fit)

km.fit2 <- survfit(Surv(InfectionTime, Infection) ~ 1, data = survival.dat)

ggsurvplot(km.fit2, data = survival.dat, pval.method = TRUE, conf.int = TRUE, censor = F, risk.table = T,tables.height = 0.3)

print(km.fit2)


```

